Smart Representations: {R}ationality, Regret, Evolution \& the Environment

### 2015-11-05 paper/paper.tex
rewrite up to Sec 3.3

\title{Rationality, Regret, Evolution \& the Environment}
\author{Paolo Galeazzi \& Michael Franke}
\date{current version: November 4, 2015}

@@

Evolutionary game theory has become an established philosophical tool for probing into
conceptual issues whose complexity requires mathematical modeling. However, most applications
of EGT look at evolutionary selection of behavior in a single \emph{stage game}. There are
actually two crude but common simplifications in this approach. Firstly, the stage game is
usually considered fixed, closed and immutable. Secondly, what evolves are behavioral
strategies for the stage game. It has consequently been argued that, instead of focusing on a
fixed, immutable stage game and the behavioral selection for that stage game alone, we should
rather study the evolutionary selection of \emph{choice mechanisms} that are adapted in a rich
and variable environment
\citep[e.g.][]{FawcettHamblin2013:Exposing-the-be,McNamara2013:Towards-a-Riche}. % However,
% formal accounts of the adaptive value of choice mechanisms have focused, for the most part, on
% the evolutionary benefit of learning rules
% \citep[e.g.][]{ZollmanSmead2010:Plasticity-and-,SmeadZollman2013:The-Stability-o}. \todo{More
%   references Harley 1981!!!, O Conor \dots} Moreover,
But only little attention has been paid to modeling the evolutionary competition between choice
mechanisms in a game-theoretic framework \citep[but
see][]{ZollmanSmead2010:Plasticity-and-,SmeadZollman2013:The-Stability-o}.\todo{More references
  Harley 1981!!!, O Conor \dots}.

This paper makes two contributions to the literature on evolution of choice mechanisms: one
methodological and one conceptual. Firstly, on the methodological side, we introduce a
conservative extension of standard EGT that allows us to reuse notions such as evolutionary
stability / evolutionary dynamics for the study of the evolutionary competition between choice
mechanisms in a statistically variable environment. Concretely, we look at the evolutionary
competition of choice mechanisms in a ``meta-game'' whose payoffs represent the average
expected fitness choice mechanisms when playing arbitrary games (from a given class) against
themselves. In this sense, a meta-game captures (the modeller's) assumptions about the relevant
statistics of the environment in which evolutionary forces operate.

Secondly, and in order to show that our method is useful for a conceptual debate about
evolution of choice mechanisms, we argue that paying attention to the interaction of subjective
belief and subjective preference representations is important. Since almost any reasonable
learning or decision-making process will make use of some (subjective, agent-internal (possibly
unconscious)) representation of the quality of choice options, the question of which such
representations are ecologically valuable (i.e., lead to high fitness) is central. Na\"ive
intuition might convince us that preferences that veridically represent the actual, true
fitness measure that evolutionary selection is based upon must surely be the best. We
demonstrate that, if we take into account how agents represent uncertainty, this na\"ive
conclusion is premature. Under certain reasonable epistemic conditions, non-veridical
preference representations in terms of regret can be evolutionarily beneficial.

@@

We would like to put this latter idea to the test. We present a formal argument intended to
support the conclusion that subjective preferences that differ from objective fitness can be
adaptive. We argue that it is important to take into account the potential mutual dependence of
subjective preferences and subjective beliefs in adaptationist approaches to decision
making. More specifically, we show that subjective, non-veridical preference representations
based on a notion of regret
\citep[e.g.][]{Savage1951:The-theory-of-s,LoomesSugden1982:Regret-Theory:-}\todo{add
  references?}\ can be adaptive and even outperform objective, veridical representations of the
actual, true fitness, when agents, at least on occasion, hold imprecise probabilistic beliefs (CITE)
and make decisions based on a security choice rule (here: the maximin rule).

The method by which we want to advance our argument is to integrate ideas from the recent
literature on preference evolution (CITE) with basic tenets of adaptationism: we are interested
in studying the evolutionary competition of choice mechanisms, conceived here as pairs of
subjective preference and subjective belief representations. Evolutionary competition between
choice mechanisms is driven by the expected objective fitness that each choice mechanisms
accrues when put to use repeatedly in a range of choice situations. We focus on the more
complex case of multi-agent decision making first, i.e., game situations. Solitary decision
making is then easy to deal with subsequently. The next section introduces the choice
mechanisms we focus on. After that, Section~\ref{sec:model} introduces the ``meta-game''
modeling approach.

@@

With some due simplification, there are two types of approaches to solving games. One looks at
a collection of players (e.g., a whole population of them) and considers sets of choices of all
players that are, in some sense or other, stable or in equilibrium. Another approach is to
focus on each individual agent's perpective and specify which of her actions the agent
\emph{should} choose if she is rational. We will look at the latter here; the former will be
relevant only later, when turn to evolutionary competition between choice mechanisms in
Section~\ref{sec:model}. We will first look at what we call \emph{choice rules}: classical
proposals of how rational agents should choose acts. We will then introduce our notion of
\emph{choice mechanism} which makes explicit several elements that choice rules might brush
under the table.

@@

For the purposes of this paper, we think of a choice rule as a function that takes a
(strategic) game as input and delivers a choice of action for that game. The literatures has
suggested a number of choice rules that aspire to capture a normative element of rational
choice. For reasons that will become clear as we go along, we focus here on three classical
choice rules: (i) maximin, (ii) Laplace and (iii) regret minimization.

But first an important digression. Rational choice from maximization of expected utilities, as
given in Definition~\ref{def:rationality}, is \emph{not} a choice rule in the present sense, if
we apply it strictly. Although the notion applies to game situations, of course, it requires an
additional component, namely a belief of the DM, before it can say which acts the DM should
choose. Let $A_{-i} = \bigtimes A_j$ for all $j \neq i$ be the set of all \emph{profiles} of
acts for all players except $i$. If player $i$ has a belief $\mu_i \in \Delta(A_{-i})$ about
the behavior of his opponents, he chooses rationally only if his choice $a^*_i$ maximizes
expected utility
$\EU(a) = \sum_{\act_{-i} \in \Acts_{-i}} \mu_i(\act_{-i}) \myts \pi_i(a_i, a_{-i})$. So this
requires a probabilistic belief about the opponents' action choices. In the example from
Figure~\ref{coordgame1}, a rational player would choose act $I$ exactly when he believes that
his opponent will choose act $I$ with a probability at least $\nicefrac{2}{3}$.

@@

The final choice rule that is relevant for our purposes here is \emph{regret
  minimization}. Regret theory has been introduced in decision theory by
\citet{Savage1951:The-theory-of-s} and Niherens [], and later developed by Bell [1982], Loomes
and Sugden [1982], and Fishburn [1982] independently. Lately,
\citet{HalpernPass2012:Iterated-Regret} showed how the use of regret minimization can give
solutions to game theoretical puzzles (like the Traveller's dilemma) in a way that is closer to
everyday intuition and empirical data. In this paper the notion of regret is defined as in
\citet{HalpernPass2012:Iterated-Regret}:

@@

To illustrate, Figure~\ref{coordgame1reg} shows the pairwise regrets and the regrets of each
row-player action for the coordination game in Figure \ref{coordgame1}. We have
$\text{reg}(I)=2$, because when the other player chooses action $I$ the pairwise regret
$reg(I,I)$ is $0$ since $I$ is a best reply to $I$, and when the other player chooses $II$ the
pairwise regret $reg(I, II)$ is $2$ since $\pi_i(II,II)-\pi_i(I,II)=2$. With the same
reasoning, $\text{reg}(II)=1$. A regret minimizing player would therefore choose act $II$.

@@

  \begin{subfigure}[b]{0.3\textwidth}
    \centering
    \begin{tabular}{ccc}
      \toprule
      & I & II \\
      \midrule
      I & 1 & 0 \\
      II & 0 & 2\\
      \bottomrule
    \end{tabular}
    \caption{Coordination game $G^C$}
    \label{coordgame1}
  \end{subfigure}
  \hspace{1cm}
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \begin{tabular}{cccc}
      \toprule
      & $reg( \cdot, I)$ & $reg(\cdot, II)$ & $reg(\cdot)$ \\
      \midrule
      I  & 0 & 2 & 2 \\ 
      II & 1 & 0 & 1\\
      \bottomrule
    \end{tabular}
    \caption{Pairwise regrets representation $\tau^{reg}(G^C)$}
    \label{coordgame1reg}
  \end{subfigure}
  \caption{A coordination game (left) and regrets associated with each row-player actions (right).}
    \label{coordgame1mainFig}
\end{figure}

@@

We think of a choice mechanism as a function that maps a game (or more generally any decision
situation) onto a triple that consists of: (i) the DM's \emph{subjective preferences} (possibly
diverging from the actual objective payoffs defined ) in the game, (ii) the DM's
\emph{subjective beliefs} about the opponents' behavior, and (iii) an \emph{action selection
  function} which maps the former two components onto a concrete choice of action in the
game. A choice mechanism could be a set of distinct heuristics (the DM's \emph{adaptive
  toolbox}) that have little in common
\citep[e.g.][]{TverskyKahnemann1981:The-Framing-of-,GigerenzerGoldstein1996:Reasoning-the-F,ScheibehenneRieskamp2013:Testing-the-Ada}. But
to keep matters simple, we will here only look at choice mechanisms that use a uniform way of
forming preferences and beliefs and a uniform way of making choices based on them. More
concretely, we focus on two \emph{preference types} and two \emph{belief types}, and neglect
variation in the action selection component.

We are most interested here in studying whether evolutionary pressure could support
non-veridical subjective preferences under certain conditions. We therefore focus on two
\emph{preference types}: veridical types $\tau^\pi$, who use the actual payoff function $\pi$ of the
actual game $G$, and regret types $\tau^{reg}$, who look at the regret transformation of the
actual game $G$.

It turns out that whether $\tau^{reg}$ can have a fitness advantage over $\tau^\pi$ depends on
the general \emph{kind} of (probabilistic) beliefs that DM's form about the opponent's
behavior. In other words, we even abstract away from the content of the DM's beliefs about his
opponents' behavior that he could have acquired by learning or observation, but first turn on
the more general question of how uncertainty should be represented in order to interact with
subjective preferences in such a way that the resulting behavior will be fitness enhancing for
the DM in the long run. We consider therefore two \emph{belief types} (to be defined below):
imprecise belief types $\Delta(A_{-i})$, who have a set of probabilistic beliefs about the
opponents' behavior and precise belief types $\overline{\mu}$ who have a single probabilistic
belief.

The following subsection describes what these belief types are and how a single action
selection function (maximin expected utility) gives rise to an intersting parallel with
classical choice rules.

\subsection{Choice mechanisms take2}

In the previous section we introduced choice rules generating behavior in a game as functions from a game to an action choice. In the general form, given a class of games $\mathcal{G}$, a choice rule is then thought as a function $c: \mathcal{G} \rightarrow A_i $. In this section instead we want to show that a more intelligible and perspicuous way of understanding behavior generating mechanisms may be as functions of a subjective preference and an epistemic state of the agents. Then, given a set of subjective preferences $ \Theta $ [@Paolo: minimally define subjective preferences somewhere above] and a set of epistemic states (or belief types) $E$, a choice mechanism is a function $\kappa: \Theta \times E \rightarrow A_i $. We will see in a moment that from this perspective we gain a more unified and structured insight on the choice rules presented above.

More
concretely, we focus on two \emph{preference types} and two \emph{belief types}. We view a preference type as a function $\tau: \mathcal{G} \rightarrow \Theta$, and a belief type as a general disposition to form beliefs about the co-player’s behavior. The two possible dispositions that we take account of are probabilistic (precise) belief and non-probabilistic (imprecise) belief, as we will explain in the next section.

We are most interested here in studying whether evolutionary pressure could support
non-veridical subjective preferences under certain conditions. We therefore consider two
\emph{preference types}: veridical type $\tau^\pi$, who uses the actual payoff function $\pi$ of the
actual game $G$, and regret type $\tau^{reg}$, who looks at the regret transformation of the
actual game $G$.

The choice mechanism that we use in this paper is \textit{maxmin expected utility} [GS89], for reasons that will be clear and justified in the following two sections.

>>>>

\title{Smart Representations: {R}ationality, Regret, Evolution \& the Environment}
\author{Paolo Galeazzi \& Michael Franke} \date{current version: November 5, 2015}

@@

Evolutionary game theory has become an established philosophical tool for probing into
conceptual issues whose complexity requires mathematical modeling. However, most applications
look at evolutionary selection of behavior in a single \emph{stage game}. There are two common
simplifications in this approach. Firstly, the stage game is usually considered fixed, closed
and immutable. Secondly, what evolves are behavioral strategies for the stage game. It has
consequently been argued that, instead of focusing on a fixed, immutable stage game and the
behavioral selection for that stage game alone, we should rather study the evolutionary
competition of \emph{choice mechanisms} under selective pressure in a rich and variable
environment \citep[e.g.][]{FawcettHamblin2013:Exposing-the-be,McNamara2013:Towards-a-Riche}.
But only little attention has been paid to modeling the evolutionary competition between choice
mechanisms in a game-theoretic framework \citep[but
see][]{ZollmanSmead2010:Plasticity-and-,SmeadZollman2013:The-Stability-o}.\todo{More references
  Harley 1981!!!, O Conor \dots}.

If agents deal with a rich and varied environment, they face many different choice
situations. A crucial part of a choice mechanism (conceived of as a mapping from a choice
situation to a concrete choice of action) is a \emph{subjective representation} of that choice
situation, i.e., the agents' manner of representing their preferences and the manner of forming
decision relevant beliefs. Since almost any reasonable learning or decision-making process will
make use of some (subjective, agent-internal (possibly unconscious)) representation of the
quality of choice options, the question of which such representations are ecologically valuable
(i.e., lead to high fitness) is central and has been subject of recent interest
(\textcolor{red}{CITE literature on evolution of preferences}). Na\"ive intuition might
convince us that preferences that veridically represent the actual, true fitness measure that
evolutionary selection is based upon must surely be the best. We demonstrate that, if we take
into account how agents represent uncertainty, this na\"ive conclusion is premature. Under
certain reasonable epistemic conditions, non-veridical preference representations in terms of
regret can be evolutionarily beneficial.

To show this, we study the evolutionary competition of choice mechanisms, or more specifically
the competition of subjective representations of choice situations, in a variable
environment. Concretely, we look at evolutionary ``meta-games'' whose ``acts'' are types of
subjective representations of games and whose ``payoffs'' represent the average expected
fitness that choice mechanisms using these representation types would accrue when playing
arbitrary games (from a given class, with a given occurrence probability) against other
subjective representations of games. In this sense, a meta-game captures (the modeller's)
assumptions about the relevant statistics of the environment in which evolutionary forces
operate, while we are still able to use standard methods from evolutionary game theory, like
stability or evolutionary dynamics.

@@

We would like to put this latter idea to the test. We present a formal argument intended to
support the conclusion that subjective preferences that differ from objective fitness can be
adaptive. We argue that it is important to take into account the potential mutual dependence of
subjective preferences and subjective beliefs in adaptationist approaches to decision
making. In other words, we focus on the evolutionary competition between \emph{subjective
  representation schemes}: general ways of forming subjective preferences and types of
subjective beliefs. We show that subjective, non-veridical preference representations based on
a notion of regret \citep[e.g.][]{Savage1951:The-theory-of-s,LoomesSugden1982:Regret-Theory:-}
can be adaptive and even outperform objective, veridical representations of the actual, true
fitness, when agents, at least on occasion, hold \emph{imprecise probabilistic beliefs}
(\textcolor{red}{CITATIONS}) and make decisions based on a security choice rule (here: the
maximin rule applied to expected utility). The next section introduces all of these notions
(precise/imprecise beliefs, veridical/regret representations, \dots) in detail.

% The method by which we want to advance our argument is to integrate ideas from the recent
% literature on preference evolution (CITE) with basic tenets of adaptationism: we are interested
% in studying the evolutionary competition of choice mechanisms, conceived here as pairs of
% subjective preference and subjective belief representations. Evolutionary competition between
% choice mechanisms is driven by the expected objective fitness that each choice mechanisms
% accrues when put to use repeatedly in a range of choice situations. We focus on the more
% complex case of multi-agent decision making first, i.e., game situations. Solitary decision
% making is then easy to deal with subsequently. The next section introduces the choice
% mechanisms we focus on. After that, Section~\ref{sec:model} introduces the ``meta-game''
% modeling approach.

@@

With some due simplification, there are two types of approaches to solving games. One looks at
a collection of players (e.g., a whole population of them) and considers sets of choices of all
players that are, in some sense or other, stable or in equilibrium. Another approach is to
focus on each individual agent's perpective and specify which of her actions the agent
\emph{should} choose if she is rational. We will look at the latter here; the former will be
relevant only later, when turn to evolutionary competition between choice mechanisms in
Section~\ref{sec:model}. We will first look at what we call \emph{choice rules}: classical
proposals of how rational agents should choose acts. We will then introduce our notion of
\emph{choice mechanism} which makes explicit several elements that choice rules might brush
under the table, in particular the crucial notion of a subjective representation scheme whose
evolutionary selection we are most interested in.

@@

For the purposes of this paper, we think of a choice rule as a function that takes a
(strategic) game as input and outputs a choice of action for that game. The literatures has
suggested a number of choice rules that aspire to capture a normative element of rational
choice. For reasons that will become clear as we go along, we focus here on three classical
choice rules: (i) maximin, (ii) Laplace and (iii) regret minimization.

Rational choice from maximization of expected utilities, as given in
Definition~\ref{def:rationality}, is \emph{not} a choice rule in the present sense. Although
the notion applies to game situations, of course, it requires an additional component, namely a
belief of the DM, before it can say which acts the DM should choose. Let
$A_{-i} = \bigtimes A_j$ for all $j \neq i$ be the set of all \emph{profiles} of acts for all
players except $i$. If player $i$ has a belief $\mu_i \in \Delta(A_{-i})$ about the behavior of
his opponents, he chooses rationally only if his choice $a^*_i$ maximizes expected utility
$\EU(a) = \sum_{\act_{-i} \in \Acts_{-i}} \mu_i(\act_{-i}) \myts \pi_i(a_i, a_{-i})$. So this
requires a probabilistic belief about the opponents' action choices. In the example from
Figure~\ref{coordgame1}, a rational player would choose act $I$ exactly when he believes that
his opponent will choose act $I$ with a probability at least $\nicefrac{2}{3}$.

@@

The final choice rule that is relevant for our purposes here is \emph{regret
  minimization}. Regret theory has been introduced in decision theory by
\citet{Savage1951:The-theory-of-s} and Niherens [], and later developed by Bell [1982], Loomes
and Sugden [1982], and Fishburn [1982] independently.\todo{fix references}\ Lately,
\citet{HalpernPass2012:Iterated-Regret} showed how the use of regret minimization can give
solutions to game theoretical puzzles (like the Traveller's dilemma) in a way that is closer to
everyday intuition and empirical data. In this paper the notion of regret is defined as in
\citet{HalpernPass2012:Iterated-Regret}:

@@

To illustrate, Figure~\ref{coordgame1reg} shows the pairwise regrets and the regrets of each
row-player action for the coordination game in Figure \ref{coordgame1}. We have
$\text{reg}(I)=2$, because when the other player chooses act $I$ the pairwise regret
$\text{reg}(I,I)$ is $0$ since $I$ is a best reply to act $I$, and when the other player
chooses act $II$ the pairwise regret $\text{reg}(I, II)$ is $2$ since
$\pi_i(II,II)-\pi_i(I,II)=2$. With the same reasoning, $\text{reg}(II)=1$. A regret minimizing
player would therefore choose act $II$.

@@

  \begin{subfigure}[b]{0.3\textwidth}
    \centering
    \begin{tabular}{ccc}
      \toprule
      & I & II \\
      \midrule
      I & 1 & 0 \\
      II & 0 & 2\\
      \bottomrule
    \end{tabular}
    \caption{Coordination game $G^C$}
    \label{coordgame1}
  \end{subfigure}
  \hspace{1cm}
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \begin{tabular}{cccc}
      \toprule
      & $\text{reg}( \cdot, I)$ & $\text{reg}(\cdot, II)$ & $\text{reg}(\cdot)$ \\
      \midrule
      I  & 0 & 2 & 2 \\ 
      II & 1 & 0 & 1\\
      \bottomrule
    \end{tabular}
    \caption{Pairwise regrets and regrets}
    \label{coordgame1reg}
  \end{subfigure}
  \caption{A coordination game (left) and regrets associated with each row-player acts (right).}
    \label{coordgame1mainFig}
\end{figure}

@@

A \emph{choice mechanism} is a map from choice situations to action choices that, unlike choice
rules, contains an explicit level of subjective representation of a decision situation. For our
purposes, we should think of a choice mechanism is a pair of functions $\gamma$ and $\kappa$,
where $\gamma$ is a \emph{subjective representation scheme} and $\kappa$ is an \emph{action
  selection function} (see FigureXYZ).\todo{make figure when notation and terminology are
  fixed}\ The subjective representation scheme $\gamma$ maps a game (or more generally any
decision situation) onto a subjective representation of the game (or choice situation) and
other decision-relevant beliefs (e.g., about what the opponent is likely to play). The action
selection function $\kappa$ takes such a subjective representation that $\gamma$ returns, and
maps it onto a concrete action choice. The crucial bit for our purposes here is the
representation function $\gamma$, because we are interested in which ways of conceptualizing a
decision situation are evolutionarily successful.

More generally even, a choice mechanism could be a set of distinct heuristics (the DM's
\emph{adaptive toolbox}) that have little in common
\citep[e.g.][]{TverskyKahnemann1981:The-Framing-of-,GigerenzerGoldstein1996:Reasoning-the-F,ScheibehenneRieskamp2013:Testing-the-Ada}. But
to keep matters simple and implementable for a start, we will here only look at choice
mechanisms that differ only with respect to their subjective representation function $\gamma$,
which in turn is conceived of as a \emph{uniform} way of forming preference and belief
\emph{types}. \textcolor{red}{Formally, $\gamma$ is a function in
  $\mathcal{G} \rightarrow \Theta \times E$, where \dots. Then, given a set of subjective
  preferences $ \Theta $ [@Paolo: minimally define subjective preferences somewhere above] and
  a set of epistemic states (or belief types) $E$, a choice mechanism is a function
  $\kappa: \Theta \times E \rightarrow A_i $.}

\textcolor{red}{More concretely, we focus on two \emph{preference types} and two \emph{belief
    types}. We view a preference type as a function $\tau: \mathcal{G} \rightarrow \Theta$, and
  a belief type as a general disposition to form beliefs about the co-player’s behavior. The
  two possible dispositions that we take account of are probabilistic (precise) belief and
  non-probabilistic (imprecise) belief, as we will explain in the next section.}

\textcolor{red}{We are most interested here in studying whether evolutionary pressure could
  support non-veridical subjective preferences under certain conditions. We therefore consider
  two \emph{preference types}: veridical type $\tau^\pi$, who uses the actual payoff function
  $\pi$ of the actual game $G$, and regret type $\tau^{reg}$, who looks at the regret
  transformation of the actual game $G$.}

\textcolor{red}{The choice mechanism that we use in this paper is \textit{maxmin expected
    utility} [GS89], for reasons that will be clear and justified in the following two
  sections.  }

% It turns out that whether $\tau^{reg}$ can have a fitness advantage over $\tau^\pi$ depends on
% the general \emph{kind} of (probabilistic) beliefs that DM's form about the opponent's
% behavior. In other words, we even abstract away from the content of the DM's beliefs about his
% opponents' behavior that he could have acquired by learning or observation, but first turn on
% the more general question of how uncertainty should be represented in order to interact with
% subjective preferences in such a way that the resulting behavior will be fitness enhancing for
% the DM in the long run. We consider therefore two \emph{belief types} (to be defined below):
% imprecise belief types $\Delta(A_{-i})$, who have a set of probabilistic beliefs about the
% opponents' behavior and precise belief types $\overline{\mu}$ who have a single probabilistic
% belief.

% The following subsection describes what these belief types are and how a single action
% selection function (maximin expected utility) gives rise to an intersting parallel with
% classical choice rules.
